A Type System for Structured Aliasing in Parallel Programs

### 2012-10-30 doc/types-pldi2013/intro.tex
More edits

To enforce this restriction, our type system relies on a combination of static 
and dynamic analyses.  The key observation behind our type system is that dynamic
analysis is very easy and relatively cheap when done at the granularity of
regions instead of individual heap locations.  To that end, our type system supports
static verification of all pointer dereferences and passing of region privileges at 
task calls boundaries.  The Legion runtime then leverages its knowledge of logical
region aliasing to perform dynamic safety checks for the unpacking
and packing of logical regions to/from heap with regards to task privileges.  
These checks are then amortized over all the uses of the logical region 
which significantly reduces their cost.

To enable parallelism, Legion again relies on a combination of static and dynamic
analysis to prove that two tasks operate on independent sets of data.  If two tasks
cannot be proven independent statically, the Legion runtime performs dynamic dependence
tests to detect region aliasing.  Since these tests are performed at the granularity of
regions instead of individual heap locations, they are significantly less costly.
In addition to supporting parallel execution for independent tasks, Legion also enables
programmer specified parallel execution in the presence of aliasing.  For example, two tasks 
accessing aliased data may only need to be serialized
with respect to each other (e.g. run atomically) or are permitted to run
simultaneously with the programmer using separate synchronization mechanisms
to control access (e.g. locks).  To support these relaxed access modes
Legion provides {\em cohrence modes} on logical regions.  For each region
argument to a task, a programmer specifies the coherence in which other tasks
can access aliased data while the task is running.  We present an operational semantics
for coherence modes and show that it is sound.

>>>>

To enforce this restriction, our type system relies on a combination of static 
and dynamic analyses.  The key observation is that dynamic
analysis is very easy and relatively cheap when done at the granularity of
regions instead of individual heap locations.  To that end, our type system supports
static verification of all pointer dereferences and passing of region privileges at 
task calls boundaries.  The Legion runtime then leverages its knowledge of logical
region aliasing to perform dynamic safety checks for the unpacking
and packing of logical regions to/from heap with regards to task privileges.  
These checks are then amortized over all the uses of the logical region 
which significantly reduces their cost.

To enable parallelism, Legion again relies on a combination of static and dynamic
analysis to prove that two tasks operate on independent sets of data.  If two tasks
cannot be proven independent statically, the Legion runtime performs dynamic dependence
tests to detect region aliasing.  Since these tests are performed at the granularity of
regions instead of individual heap locations, they are significantly less costly.
In addition to supporting parallel execution for independent tasks, Legion also enables
programmer specified parallel execution in the presence of aliasing.  For example, two tasks 
accessing aliased data may only need to be serialized
with respect to each other (e.g. run atomically) or are permitted to run
simultaneously with the programmer using separate synchronization mechanisms
to control access (e.g. locks).  To support these relaxed access paradigms 
Legion provides {\em cohrence modes} on logical regions.  For each region
argument to a task, a programmer specifies the coherence in which other tasks
can access aliased data while the task is running.  We present an operational semantics
for coherence modes and show that it is sound.

### 2012-10-30 doc/types-pldi2013/langdef.tex
minor edits
untangled the rules figure
More edits

\begin{figure*}
\centering{
$
\begin{array}{@{}l@{}}
\begin{array}{@{}l@{\hspace{-0.25in}}l@{}} 
\begin{array}{@{}lll}
apply(S, []) & = & S \\
apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E) \\
apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \\
& & \text{ where } S' = apply(S, E)
\end{array} &
\begin{array}{@{}ll}
mark\_coh([], \hat Q) & = [] \\
mark\_coh([ op(l, c, v) ] \dplus E, \hat Q) & = [ op(l, c', v) ] \dplus mark\_coh(E, \hat Q), \\
\multicolumn{2}{@{}l}{ \hspace{1.0in} \text{ where } c' =
\begin{cases}
simult, & \text{if }\exists \rho. l \in \rho \wedge simult(\rho) \in \hat Q \\
atomic, & \text{if }\exists \rho. l \in \rho \wedge atomic(\rho) \in \hat Q \\
excl, & \text{otherwise}
\end{cases}}
\end{array}
\end{array} \\
\begin{array}{@{}lll}
any\_interleave([], [], \ldots, []) & = & true \\
any\_interleave([ \epsilon ] \dplus E', E_1, \ldots, [ \epsilon ] \dplus E_i, \ldots, E_n) & = & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
valid\_interleave(S, C, E', \oton{E}) & \Rightarrow & any\_interleave(E', \oton{E}) \\
\end{array}
\end{array}
$}
\caption{Operational Semantics Helper Functions}
\label{fig:opsemfns}
\vspace{-5mm}
\end{figure*}

@@

Theorem~\ref{thm:effects} guarantees that a function call produces a value of the
correct type and does so within the regions of memory for which privileges are held.  In fact, this results holds
even if the {\em valid\_interleave} constraint is weakened to only require an arbitrary interleaving of the memory
traces of subexpressions (i.e. the {\em any\_interleave} predicated defined in Figure~\ref{fig:opsecfns}).  The 
additional constraints on {\em valid\_interleave} in Figure~\ref{fig:validinterleave} address the coherence of 
heap accesses and allow for the per-region relaxation of coherence that is introduced in Legion to enable parallelism
in the presence of aliasing.

To determine whether an interleaving of two or more memory traces is valid, we consider three sets of addresses:
\begin{itemize}
\item exclusive locations ($l \in L_{excl}$) are those which have at least once access in exclusive mode in the traces
and are not in the clobber set.  For these locations, we require the standard {\em sequential consistency} - all
reads to these locations see the effect of previous writes and reductions, and the resulting state of the store is
as if all writes and reductions were applied from each trace in order
\item atomic locations ($l \in L_{atomic}$) are those which have at least one access in atomic mode in the traces and
are in neither $L_{excl}$ nor the clobber set.  For these locations, we allow permutations of the original
subexpression trace order.
\item for locations that have only access in {\em simult} mode or are in the clobber set, no constraints are
enforced.  The validity of the interleaving of these accesses will be performed in the an enclosing expression.
\end{itemize}

@@

Serial execution is achieved by ignoring the coherence mode $Q$ in all function calls, using $Q' = \emptyset$
instead, and by forming ``interleaved'' traces by simply concatenating the subexpressions' traces.  By ignoring
the coherence modes, the clobber set remains empty, making the result of all {\em read} expressions fully
determined.  The following lemma and theorem (proofs of which can be found in \cite{LegionTypes12}) show that the
value and memory trace that result from a serialized execution of an expression are always valid executions of
Legion programs.

@@

To determine when parallel execution is safe, we start from the serial execution trace and allow the reordering of 
adjacent heap operations that do not change the behavior of the application.  If we can show that reordering is allowed
between any two operations that come from different constituent traces, any interleaving of the constituent traces will
be equivalent to the serialized execution and parallel execution was safe.  Finally, to allow the safety of parallel
execution to be determined {\em before} the execution is actually performed, we show that a test based on the subtask
privileges and the current region mapping can soundly predict when this property will hold.

@@

The final piece is to turn this test into something that can be performed before the subtasks are executed,
ideally as efficiently as possible.  The key to this is the privileges on Legion tasks and their ability to
provide bounds on the accesses that may be performed during execution.  Our non-interference operator is
extended to work on privileges and sets of privileges:
$
\begin{array}{l@{}l}
\multicolumn{2}{l}{priv_1(r_1) \nonint[M] priv_2(r_2) \Leftrightarrow} \\
& (priv_1 = reads \wedge priv_2 = reads)\ \vee \\
& (priv_1 = reduces_{id_1} \wedge priv_2 = reduces_{id_2} \wedge id_1 = id_2)\ \vee \\
& (r_1 * r_2)\ \vee \\
& (M(r_1) \cap M(r_2) = \emptyset) \\
\\
\multicolumn{2}{l}{\Phi_1 \nonint[M] \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \nonint[M] \phi_2}
\end{array}$
The ``two readers'' and ``two reducers'' conditions for memory operation non-interference have exact equivalents
for regions, which can be performed at compile-time.  The ``different address'' condition is approximated by two
tests.  The first uses (logical) region disjointness constraints from the type system to statically infer
non-interference.  The second uses the region mapping $M$ to determine
the two physical regions and performs a run-time test for overlapping.  Although still a dynamic test, it is performed
once per pair of regions rather than for for every pair of memory operations, and often need not be performed at all
(if one of the other three conditions is shown to hold at compile-time).

>>>>

\begin{figure*}
\centering{
$
\begin{array}{@{}l@{}}
\begin{array}{@{}l@{\hspace{-0.25in}}l@{}} 
\begin{array}{@{}lll}
apply(S, []) & = & S \\
apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E) \\
apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \\
& & \text{ where } S' = apply(S, E)
\end{array} &
\begin{array}{@{}ll}
mark\_coh([], \hat Q) & = [] \\
mark\_coh([ op(l, c, v) ] \dplus E, \hat Q) & = [ op(l, c', v) ] \dplus mark\_coh(E, \hat Q), \\
\multicolumn{2}{@{}l}{ \hspace{1.0in} \text{ where } c' =
\begin{cases}
simult, & \text{if }\exists \rho. l \in \rho \wedge simult(\rho) \in \hat Q \\
atomic, & \text{if }\exists \rho. l \in \rho \wedge atomic(\rho) \in \hat Q \\
excl, & \text{otherwise}
\end{cases}}
\end{array}
\end{array} \\
\begin{array}{@{}lll}
any\_interleave([], [], \ldots, []) & = & true \\
any\_interleave([ \epsilon ] \dplus E', E_1, \ldots, [ \epsilon ] \dplus E_i, \ldots, E_n) & = & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
valid\_interleave(S, C, E', \oton{E}) & \Rightarrow & any\_interleave(E', \oton{E}) \\
\end{array}
\end{array}
$}
\caption{Operational Semantics Helper Functions \label{fig:opsemfns}}
\vspace{-5mm}
\end{figure*}

@@

An important insight in the proof of Theorem~\ref{thm:effects} was that 
it did not rely on the definition of {\em valid\_interleave}.  Weakening 
{\em valid\_interleave} to {\em any\_interleave} (Figure~\ref{fig:opsemfns}), 
which only requires any arbitrary interleaving of memory
traces from subexpressions, 
would have been a sufficient condition for the proof.  
The additional constraints in {\em valid\_interleave}
(Figure~\ref{fig:validinterleave}) address the {\em coherence} of heap accesses.
Coherence defines the permissible ordering of operations to each individual heap 
location by parallel computations.  The constraints enforced by {\em valid\_interleave}
specify exactly what interleavings of memory traces are permitted based on
the per-region coherence modes introduced by Legion for supporting parallelism
in the presence of aliasing.

To determine whether an interleaving of two or more memory traces is valid, we consider three sets of addresses:
\begin{itemize}
\item exclusive locations ($l \in L_{excl}$) are those which have at least one access in exclusive mode in the traces
and are not in the clobber set.  For these locations, we require the standard {\em sequential consistency} - all
reads to these locations see the effect of previous writes and reductions, and the resulting state of the store is
as if all writes and reductions were applied from each trace in order.
\item atomic locations ($l \in L_{atomic}$) are those which have at least one access in atomic mode in the traces and
are in neither $L_{excl}$ nor the clobber set.  For these locations, we allow permutations of the original
subexpression trace order.
\item for locations that have only access in {\em simult} mode or are in the clobber set, no constraints are
enforced.  The validity of the interleaving of these accesses will be performed in the an enclosing expression.
\end{itemize}

@@

Serial execution is achieved by ignoring the coherence mode $Q$ in all function calls, using $Q' = \emptyset$
instead, and simply interleaving traces by concatenating the subexpressions' traces in program order.  By ignoring
the coherence modes, the clobber set remains empty, making the result of all {\em read} expressions fully
determined.  The following lemma and theorem (proofs of which can be found in \cite{LegionTypes12}) show that the
value and memory trace that result from a serialized execution of an expression are always valid executions of
Legion programs.

@@

To determine when parallel execution is safe, we start from the serial execution trace and allow the reordering of 
adjacent heap operations that do not change the behavior of the application.  If we can show that it is safe to 
reorder any pair of operations that come from two different constituent traces, then any interleaving of the constituent
traces will be equivalent to a sequential execution, thereby enabling parallel execution.  To
efficiently implement this requirement at runtime, we require a test that can determine this property prior
to the actual execution of the tasks that create the traces.  We show that a test based on the subtask privileges and current region 
mapping can soundly predict when this property will hold.
%If we can show that reordering is allowed
%between any two operations that come from different constituent traces, any interleaving of the constituent traces will
%be equivalent to the serialized execution and parallel execution was safe.  Finally, to allow the safety of parallel
%execution to be determined {\em before} the execution is actually performed, we show that a test based on the subtask
%privileges and the current region mapping can soundly predict when this property will hold.

@@

The final piece is to turn this test into something that can be performed before the subtasks are executed,
ideally as efficiently as possible.  The key to this is the privileges on Legion tasks and their ability to
provide bounds on the accesses that may be performed during execution.  Our non-interference operator is
extended to work on privileges and sets of privileges:
$
\begin{array}{l@{}l}
\multicolumn{2}{l}{priv_1(r_1) \nonint[M] priv_2(r_2) \Leftrightarrow} \\
& (priv_1 = reads \wedge priv_2 = reads)\ \vee \\
& (priv_1 = reduces_{id_1} \wedge priv_2 = reduces_{id_2} \wedge id_1 = id_2)\ \vee \\
& (r_1 * r_2)\ \vee \\
& (M(r_1) \cap M(r_2) = \emptyset) \\
\\
\multicolumn{2}{l}{\Phi_1 \nonint[M] \Phi_2 \Leftrightarrow \displaystyle\bigwedge_{\phi_1 \in \Phi_1, \phi_2 \in \Phi_2} \phi_1 \nonint[M] \phi_2}
\end{array}$
The ``two readers'' and ``two reducers'' conditions for memory operation non-interference have exact equivalents
for regions, which can be performed statically.  The ``different address'' condition is approximated by two
tests.  The first uses (logical) region disjointness constraints from the type system to statically infer
non-interference.  The second uses the region mapping $M$ to determine
the two physical regions and performs a run-time test for overlapping.  Although still a dynamic test, it is performed
once per pair of regions rather than for for every pair of memory operations.  Details on how to perform
this test efficiently at runtime can be found in \cite{Legion12}.
%and often need not be performed at all
%(if one of the other three conditions is shown to hold at compile-time).
